---
title: "Estimating 5-year lung cancer risk with LLPv3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating 5-year lung cancer risk with LLPv3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(LLPv3)
```

## What this package does

`LLPv3` implements version 3 of the Liverpool Lung Project risk model
(Field et al., *Thorax* 2021). Given a person's age, sex, smoking duration,
family history of lung cancer, prior pneumonia, asbestos exposure, and prior
(non-lung) cancer, it returns the **5-year probability of lung cancer**.

It reproduces the official Liverpool calculator exactly. Two interfaces are
provided: the vectorised `llpv3()` function and the ModelsCloud prediction API
(`model_run()` / `get_sample_input()` / `get_default_input()`).

## Scoring one person

```{r}
llpv3(age = 65, sex = "male", smoking_duration = 45,
      family_hist_lung_cancer = 0, pneumonia = 1,
      asbestos = 0, prior_cancer = 0)
```

Additional risk factors increase the estimate:

```{r}
llpv3(age = 72, sex = "male", smoking_duration = 55,
      family_hist_lung_cancer = 1, pneumonia = 1,
      asbestos = 1, prior_cancer = 1)
```

## How the score is built

The linear predictor is an age/sex baseline plus a sum of risk-factor log-odds.
Two effects worth noting:

* **Smoking duration** enters in bands (1–19, 20–39, 40–59, ≥60 years), not
  per-year. A never-smoker (`smoking_duration = 0`) contributes nothing from
  smoking.
* **Family history** distinguishes *early* onset (a relative diagnosed before
  60) from *late* onset (≥60). Early onset carries a larger coefficient.

```{r}
# Same person, varying only family history
fh <- c(none = 0, early = 1, late = 2)
vapply(fh, function(f)
  llpv3(age = 65, sex = "female", smoking_duration = 40,
        family_hist_lung_cancer = f, pneumonia = 0,
        asbestos = 0, prior_cancer = 0),
  numeric(1))
```

## Scoring a cohort

```{r}
cohort <- data.frame(
  age              = c(58, 64, 72),
  sex              = c("male", "female", "male"),
  smoking_duration = c(30, 44, 55),
  fam              = c(0, 1, 2),
  pneumonia        = c(0, 1, 1),
  asbestos         = c(0, 0, 1),
  prior_cancer     = c(0, 0, 1)
)
cohort$risk <- llpv3(
  age = cohort$age, sex = cohort$sex,
  smoking_duration = cohort$smoking_duration,
  family_hist_lung_cancer = cohort$fam,
  pneumonia = cohort$pneumonia, asbestos = cohort$asbestos,
  prior_cancer = cohort$prior_cancer
)
cohort[, c("age", "sex", "risk")]
```

## Using the ModelsCloud API

```{r}
get_default_input()

person <- get_default_input()
person$age <- 70
person$smoking_duration <- 50
model_run(person)
```

```{r}
model_run(get_sample_input())
```

## Caveats

* Developed in a Caucasian population; intended for ages 50–79.
* For research use only; differs from LLPv2 used in UK screening programmes.
* Risk is a probability — eligibility thresholds are a separate decision and are
  not applied here.

## Reference

Field JK, Vulkan D, Davies MPA, Duffy SW, Gabe R. Liverpool Lung Project lung
cancer risk stratification model: calibration and prospective validation.
*Thorax.* 2021;76(2):161–168.
doi:[10.1136/thoraxjnl-2020-215158](https://doi.org/10.1136/thoraxjnl-2020-215158)
